\documentclass[a4paper,oneside]{book}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphics}         % used to include figures
\usepackage{picins}           % permits to include figures into text
\usepackage{times}
\usepackage{mathtime}
\usepackage{url}

%\pdfoutput=1                  % pdflatex outputs pdf by default
%\pdfcompresslevel=0           % no compression

\makeatletter
% Good subscript
\DeclareRobustCommand*\down[1]{%
   \@textsubscript{\selectfont#1}}
\def\@textsubscript#1{%
   {\m@th\ensuremath{_{\mbox{\fontsize\sf@size\z@#1}}}}}
\makeatother

\includeonly{introduction,conffiles,spatial,tkstochsim,gillespie}

\newcommand{\stochsim}{S\textsc{toch}S\textsc{im}}

\title{\stochsim{}, the stochastic simulator}
\author{Carl Firth, Nicolas Le Novère and Tom Shimizu}

% Here are the macros describing the quantitative limits of the program 
\newcommand{\CURRENTVERSION}{1.4} 
\newcommand{\MAXCOMPONENTNAMELENGTH}{32} 
\newcommand{\MAXCOMPONENTSYMBOLLENGTH}{8}  
\newcommand{\MAXCOMPONENTS}{128} 
\newcommand{\MAXCOMPLEXTYPES}{128} 
\newcommand{\MAXCOMPONENTSINCOMPLEX}{64} 
\newcommand{\MAXLIMITCOMPLEXES}{16777216}
\newcommand{\MAXNUMDISPLAYVARIABLES}{128} 
\newcommand{\MAXNUMREACTIONSPERCOMPLEX}{64} 
\newcommand{\MAXOUTPUTVARNAMELENGTH}{64} 
\newcommand{\MAXTYPESINOUTPUTVAR}{16}
\newcommand{\MAXMSFLAGLENGTH}{8} 
\newcommand{\MAXMSNUMFLAGS}{31}
\newcommand{\MAXNUMCOMPLEXARRAYS}{8} 
\newcommand{\MAXNUMARRAYSNAPSVARIABLES}{128}
\newcommand{\MAXNUMARRAYSNAPSSTATES}{16}
\newcommand{\MAXARRAYWIDTH}{128}
\newcommand{\MAXARRAYLENGTH}{128}
\newcommand{\MAXARRAYNAMELENGTH}{128}
\newcommand{\MAXARRAYVARNAMELENGTH}{128}

\voffset=-0.3cm
\textheight=22.5cm
\hoffset=-1cm
\textwidth=14.4cm

\begin{document}
\maketitle
\tableofcontents
\include{introduction}
\include{conffiles}
\include{spatial}
\include{tkstochsim}

\chapter{References}
\begin{itemize}
\item Bray, D., Levin, M. D. \& Morton-Firth, C. J. (1998).  Receptor
clustering as a cellular mechanism to control sensitivity. Nature 393,
85-88.

\item Duke, T. A. J. \& Bray, D.  (1999).  Heightened sensitivity of a
lattice of membrane receptors.  Proc. Natl. Acad. Sci. USA 96,
10104-10108.

\item Gillespie, D. T. (1976). A general method for numerically
simulating the stochastic time evolution of coupled chemical
reactions. J. Comput. Phys. 22, 403-434.

\item Gillespie, D. T. (1977). Exact stochastic simulation of coupled
chemical reactions. J. Phys. Chem. 81, 2340-2361.

\item McAdams, H. H. \& Arkin, A. (1997). Stochastic mechanisms in
gene expression. Proc. Natl. Acad. Sci .USA 94, 814-819.

\item Morton-Firth, C. J. (1998). Stochastic simulation of cell
signalling pathways. Ph.D. Thesis, University of Cambridge.

\item Morton-Firth, C. J. \& Bray, D. (1998). Predicting temporal
fluctuations in an intracellular signalling pathway. J. Theor. Biol.
192, 117-128.

\item Morton-Firth, C. J., Shimizu, T. S. \& Bray, D. (1999). A
free-energy-based stochastic simulation of the Tar receptor complex.
J. Mol. Biol.286, 1059-1074.

\item Shimizu, T.~S., Le~Nov\`{e}re, N., Levin, M.~D., Beavil, A.~J.,
  Sutton, B.~J., and Bray, D. (2000).  Molecular model of a lattice of
  signalling proteins involved in bacterial chemotaxis. Nat. Cell
  Biol. 2, 792--796.

\end{itemize}


\appendix
\include{gillespie}
  
\end{document}
